---
title: "Frazer_A7"
author: "Neil Frazer"
date: "2020-03-06"
output: 
  html_document: 
    theme: cerulean
    toc: true
    smooth_scroll: true
    toc_depth: 3
    toc_float: true
    number_sections: false
    code_folding: hide
---  

<style type="text/css">  
/* Note: CSS uses C-style commenting. */
h1.title{font-size:22px; text-align:center;}
h4.author{font-size:16px; text-align:center;}
h4.date{font-size:16px; text-align:center;}
body{ /* Normal  */ font-size: 13px}
td {  /* Table   */ font-size: 12px}
h1 { /* Header 1 */ font-size: 16px}
h2 { /* Header 2 */ font-size: 14px}
h3 { /* Header 3 */ font-size: 12px}
.math{ font-size: 10pt;}
.hi{ /* hanging indents */ 
    padding-left:22px; 
    text-indent:-22px;
}
blockquote {  
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 12px;
    border-left: 5px solid #eee;
}
code.r{ /* code */ 
       font-size: 12px;
}
pre{/*preformatted text*/ 
    font-size: 12px;
}
p.caption {/* figure captions */ 
    font-size: 1.0em;
    font-style: italic; 
} 
</style>

```{r setup, echo=FALSE}
rm(list=ls()) # clean up
library(knitr)
library(coda)
gr <- (1+sqrt(5))/2 # golden ratio, for figures
opts_chunk$set(comment="  ",
               eval.after="fig.cap",
               collapse=TRUE, 
               dev="png",
               fig.width=7,
               out.width="95%",
               fig.asp=0.9/gr,
               fig.align="center",
               cache=TRUE,
               autodep=TRUE
               )
source("sourceMe_A7.R")
```  

# Introduction  

In Chapter 7 we move from basic Bayesian concepts to the foundations of Markov Chain Monte Carlo (MCMC). Pages 154—156 are optional but I hope you will study them anyway, as they are one of the most accessible treatments I have seen. In them Kruschke shows that the target PDF is an equilibrium point of the Metropolis procedure. His demonstration does _not_ show that it is a _stable_ equilibrium; in theory it could be an _unstable_ equilibrium point, like a ball perched on a mountain top, but in practice such problems are vanishingly rare.  

**Note** that this assignment comes with a file `sourceMe_A7.R` containing Kruschke's `HDIofMCMC()`. The file is sourced in the setup chunk; just put it in the same folder as your A7. I am not asking you to understand how `HDIofMCMC()` works, but try to do so anyway; it's a useful function that you can apply to any set of samples.  

# Tips for figures    
Here are a couple of tricks I sometimes use when making plots. Neither is essential.   

## Captions  
To make a caption with inserted values, put `eval.after="fig.cap"` (notice the quotes) and `fig.cap=fig.cap` (notice the lack of quotes) in the chunk header, and then create `fig.cap` as a string at the end of the chunk using `paste()`. The alternative is to use the narrative to make your caption with inline code for the variable values. I often make `eval.after="fig.cap"` a default chunk option in my setup chunk.    

## legend( )  
To pack extra information into a legend, use its `title` argument. Create it using `paste()` or `paste0()`, and embed `\n` to make it multiline. When you play these kind of games, the box around the legend can cause trouble; to get rid of the box put `bty="n"` in your `legend()` call. Using `legend()` like this isn't especially pretty, but it is quicker to code than annotating with `text()`.      

```{r experiments, eval=FALSE, include=FALSE}
## Find a bimodal likelihood for this exercise
f1 <- function(x) {
  c1 <- 0.7
  c2 <- 1 - c1
  a1 <- 10
  b1 <- 20
  a2 <- 20
  b2 <- 10
  print(c(c1/beta(a1,b1), c2/beta(a2,b2)))
  7*beta(a1,b1)*dbeta(x, a1, b1) + 3*beta(a2,b2)*dbeta(x, a2, b2)
}
plot(f1, xlim=c(0,1), ylab="f(x)")
```   

# Exercise 7.4  
(Not in text) (25 points)  
In this exercise you write a Metropolis algorithm to solve a 1-D problem that is slightly more realistic than the island-hopping politician considered in Section 7.2 of our text. (Using MCMC on any 1-D problem is like using a twelve pound hammer to kill a mosquito, but we overlook that here.) Suppose we have some [compositional data](https://en.wikipedia.org/wiki/Compositional_data) with just two possible components, A and B say, indexed by $\theta$. For example, if $\theta=0.9$ the mixture is 90% A and 10% B. Suppose our likelihood function is  

$$
L(\theta\vert D) \propto \tag{1}
7\,\theta^9(1-\theta)^{19} + \ 3\,\theta^{19}(1-\theta)^9
$$
and our prior is the Haldane prior $\theta^{-1}(1-\theta)^{-1}$. (a) Use the Metropolis algorithm to generate 2,000 samples of $\theta$ from the posterior and (b) make a histogram. (c) Use `coda::effectiveSize()` to calculate the effective size of your 2,000 samples. For the proposal distribution use the normal with SD = 0.2. (d) Add the posterior PDF to your histogram as a red line. (f) Use your samples to estimate $\Pr(0.2 \le \theta \le 0.5)$. (g) Use your samples to estimate $\Pr(0.3 \le \theta \le 0.4 \, \big\vert\, 0.1 \le \theta \le 0.8)$.

Don't forget to give your plot a legend and a caption. In lieu of a legend, you may annotate your plot roughly like Kruschke does in the panels of Figure 7.1, page 146.  

# Solution 7.4  
Here is the requested histogram.    

```{r ex7.4, fig.cap=fig.cap}
par(mgp=c(2,1,0),     # move axis labels closer to axes
    mar=c(3.5,1,1,1), # reduce margins 2,3,4
    bg="grey95")      # background color
set.seed(124) # repeatability
f <- function(x) # posterior kernel
  7*x^8*(1-x)^18 + 3*x^18*(1-x)^8 

Ns <- 2000 # numer of samples  
SD <- 0.2  # sd of proposal normal distn

x    <- numeric(Ns) # theta's
post <- numeric(Ns) # f(theta's)

x[1] <- runif(1)    # starting theta
post[1] <- f(x[1])  # posterior at starting theta
for (i in 2:Ns) {
  move <- rnorm(1, mean=0, sd=SD)
  prop <- x[i-1] + move  # proposal
  
  if (prop<0 || prop>1) { # stay put
    x[i] <- x[i-1]        
    post[i] <- post[i-1]
    next
  } 
  
  fp <- f(prop)
  fn <- post[i-1]
  if (fp >= fn) { # accept
    x[i] <- prop   
    post[i] <- fp
  } else if (runif(1) < fp/fn) { # accept
    x[i] <- prop
    post[i] <- fp
  } else { # stay put
    x[i] <- x[i-1]
    post[i] <- post[i-1] 
  }
}
essx <- floor(effectiveSize(x))

## target
xx <- seq(0, 1, 0.01)
int <- integrate(f, 0, 1)$value
yy <- f(xx)/int

hist(x, freq=FALSE, breaks=seq(0, 1, 0.05), 
     ylim=c(0, 1.05*max(yy)), yaxt="n",
     xlab=expression(theta), cex.lab=1.2,
     yaxs="ir", ylab="", main="",
     border="white", col="skyblue")
## add target to plot
lines(xx, yy, col="red", lwd=2)
## add density to plot  
dens <- density(x, kernel="gaussian")
lines(dens$x, dens$y, col="black")
## add legend to plot
legTitle <- paste0("Metropolis\n",
                   "Ns = ", Ns, "\n", 
                   "ESS = ", essx, "\n",
                   "SD = ", SD)
legend("topright", bty="n", inset=c(0.1, 0.2),
       pch=c(15, NA, NA), pt.cex=c(1.8, NA, NA), 
       lty=c(NA, 1, 1), lwd=c(NA, 1, 2),
       col=c("skyblue", "black", "red"),
       title=legTitle,
       legend=c("samples", "sample density", "target")
       )
fig.cap <- paste0("**Figure 7.4.** Metropolis algorithm with a 
                  normal proposal distribution. ", Ns, 
                  " samples were used. The effective samples size was ",
                  essx, ", and the proposal standard deviation was ",
                  SD, ". See text for details of the prior and 
                  likelihood.")
## answer questions (c) and (d)
Prf <- sum(0.2 <= x & x <= 0.5)/Ns
Prg <- sum(0.3 <= x & x <= 0.4)/sum(0.1 <= x & x <= 0.8)
```  

For question (f) the answer is $\Pr(0.2 \le \theta \le 0.5)$ = `r signif(Prf*100, 3)`%, and for question (g) the answer is $\Pr(0.3 \le \theta \le 0.4 \, \big\vert\, 0.1 \le \theta \le 0.8)$ = `r signif(Prg*100, 3)`%.  

# Exercise 7.5  
(Not in text) (25 points)  
Do the same problem you did in Ex 7.4, but use the Gibbs sampler instead of the Metropolis algorithm. 

## Solution 7.5  
```{r ex7.5, fig.cap=fig.cap}
par(mgp=c(2,1,0),     # move axis labels closer to axes
    mar=c(3.5,1,1,1), # reduce margins 2,3,4
    bg="grey95")      # background color
set.seed(124)         # repeatability

f <- function(x) # posterior kernel
  7*x^8*(1-x)^18 + 3*x^18*(1-x)^8 
rf <- function(f, N=31) { # Our 1-D Gibbs sampler
  x <- seq(0,1,len=N)
  y <- f(x)
  p <- cumsum( diff(x) * 0.5 * (y[-N] + y[-1]) )
  p <- c(0, p)
  p <- p/p[N]
  u <- runif(1)
  i1 <- max(sum(p<u), 1)
  i2 <- i1 + 1
  x[i1] + (x[i2] - x[i1])*(u - p[i1])/(p[i2] - p[i1])
}

## Generate samples
Ns <- 2000 # numer of samples  
x <- numeric(Ns) # allocate storage
for (i in 1:Ns) x[i] <- rf(f=f)

essx <- floor(effectiveSize(x))
## target
xx <- seq(0, 1, 0.01)
int <- integrate(f, 0, 1)$value
yy <- f(xx)/int
## histogram
hist(x, freq=FALSE, breaks=seq(0, 1, 0.05), 
     ylim=c(0, 1.05*max(yy)), 
     xlab=expression(theta), cex.lab=1.2,
     yaxs="ir", yaxt="n", ylab="", main="",
     border="white", col="skyblue")
## add target to plot
lines(xx, yy, col="red", lwd=2)
## add density to plot  
dens <- density(x, kernel="gaussian")
lines(dens$x, dens$y, col="black")
## add legend
legTitle <- paste0("Gibbs sampler\n",
                   "Ns = ", Ns, "\n",
                   "ESS = ", essx)
legend("topright", bty="n", inset=c(0.1, 0.2),
       pch=c(15, NA, NA), pt.cex=c(1.8, NA, NA), 
       lty=c(NA, 1, 1), lwd=c(NA, 1, 2),
       col=c("skyblue", "black", "red"),
       title=legTitle,
       legend=c("samples", "sample density", "target")
       )
## make caption
fig.cap <- paste0("**Figure 7.5.** The Gibbs sampler applied to the same
                  problem as in Figure 7.4. ", "The number of samples used
                  was ", Ns, ".  The effective samples size was ",
                  essx, ". See text for details of the prior and 
                  likelihood.")
```

The sample histogram and sample density estimate the target distribution very well. It is a bit surprising that these results from the Gibbs sampler are not more noticeably superior to those from the Metropolis algorithm even though the effective sample size from the Gibbs sampler is much larger. The lesson I take from this is that it is a mistake to thin a chain (i.e., to discard correlated samples) unless storage is limited. I did not thin the Metropolis samples, but I was tempted to do so.     

# Exercise 7.6  
(Not in text) (25 points)

## Preamble  
The two-coin example of this chapter is somewhat artificial, because the two coins are _independent_. In this exercise we estimate the difference in bias between two coins using MCMC, but without resort to either Metropolis or the Gibbs sampler. As you saw in class, deriving a useful mathematical formula for the PDF of a difference $\theta_1-\theta_2$ is not an easy task. However, it is easy to _sample_ the difference, and then use those samples as a proxy for the PDF itself. 

## The task  
You have two coins that cannot communicate with each other. We refer to the probability of heads as the bias, $\theta$. Suppose the first coin has unknown bias $\theta_1$ and the second coin has unknown bias $\theta_2$. You flip the first coin $N_1=50$ times obtaining $z_1=35$ heads. You flip the second coin $N_2=40$ times, obtaining $z_1=18$ heads. Generate 2,000 samples of the difference in bias $\phi=\theta_1-\theta_2$ and use them to make a histogram. Estimate (a) the expected value of $\phi$, (b) the SD of $\phi$ and (c) a 95% HDI for $\phi$. (d) Use your samples to estimate $\Pr(-0.1 \lt \phi \lt 0.1)$. Annotate your plot with the sample mean and 95% HDI the way Kruschke does, or use `legend()`.    

# Solution 7.6    
For each coin we use the Haldane prior and take advantage of conjugacy to get the posterior PDF. We then draw 2,000 samples from each posterior and take their differences to get 2,000 samples $\phi^{(i)}=\theta_1^{(i)} - \theta_2^{(i)}$. Using these samples of $\phi$ we estimate the requested quantities.  

For the first coin the $z_1$ heads and $N_1 - z_1$ tails give a Bernoulli likelihood that when combined with the Haldane prior results in a beta posterior with $a_1 = z_1$ and $b_1=N_1 - z_1$. Similarly for the second coin. In the following chunk we make the requested samples.  

```{r ex7.6, fig.cap=fig.cap}
par(mgp=c(2.8,1,0),     # move axis labels closer to axes
    mar=c(4.2,1,1,1),   # reduce margins 2,3,4
    bg="grey95")        # background color
set.seed(124)           # repeatability
s <- function(x) signif(x, 3)
Ns <- 2000 # number of samples

z1 <- 35
N1 <- 50
a1 <- z1
b1 <- N1 - z1
theta1 <- rbeta(Ns, a1, b1) # samples

z2 <- 18
N2 <- 40
a2 <- z2
b2 <- N2 - z2
theta2 <- rbeta(Ns, a2, b2) # samples

phi <- theta1 - theta2 # samples
sampleMean <- mean(phi)
theoryMean <- z1/N1 - z2/N2

sampleVar <- var(phi)
sampleSD  <- sqrt(sampleVar)
fv <- function(a,b) # variance of theta
  a*b/(a + b)^2/(a + b + 1)
theoryVar <- # variance of phi
  fv(a1, b1) + fv(a2, b2)
theorySD  <- sqrt(theoryVar)

essPhi <- coda::effectiveSize(phi)

hist(phi, freq=FALSE, breaks=seq(-1, 1, 0.05), xlim=c(-0.2, 0.7), 
     xlab=expression(phi == theta[1] - theta[2]), cex.lab=1.2,
     yaxs="i", yaxt="n", ylab="", main="",
     border="white", col="skyblue")

CIper <- 95 # % percent credible interval
dper  <- (100 - CIper)/2
HDI95 <- HDIofMCMC(phi, CIper/100)

abline(v=c(theoryMean, sampleMean),
       lty=c(2, 1), lwd=c(2, 1), 
       col=c("red", "black"))

lines(HDI95, c(0.1, 0.1), lwd=4)
text(HDI95[1], 0.15, pos=3, cex=1.1, paste0(s(HDI95[1])))
text(HDI95[2], 0.15, pos=3, cex=1.1, paste0(s(HDI95[2])))

legend("topleft", bty="n", inset=c(0.02, 0.05),
       title="Difference in bias of two coins",
       legend=c("theory mean","sample mean", "95% HDI"),
       lty=c(2, 1, 1), lwd=c(1, 1, 4),
       col=c("red", "black", "black")
       )

fig.cap <- 
  paste0("**Figure 7.6.** Samples from the distribution of the 
          difference of two coin biases. Coin 1 had ", z1, 
         " heads in ", N2, " throws, and coin 2 had ", z2, " heads
         in ", N2, " throws. The mean from ", Ns, " samples was ",
         s(sampleMean), " whereas the mean from theory was ", s(theoryMean),
         ". The SD from samples was ", s(sampleSD), ", while
         the SD from theory was ", s(theorySD), ". The effective 
         sample size was ", essPhi, " as expected because of our method.
         The ", CIper, "% HDI is [", s(HDI95[1]),", ", s(HDI95[2]), "],
         which notably does not include zero.")

## Answer question (d)
Prd <- sum(-0.1 < phi & phi < 0.1)/Ns
```  
The answer to question (d) is $\Pr(-0.1 \lt \phi \lt 0.1)$ = `r signif(Prd*100,3)`%.  

