---
title: "A7"
author: "Michael Tucker"
date: "2020-03-06"
output: 
  html_document: 
    theme: cerulean
    toc: true
    smooth_scroll: true
    toc_depth: 3
    toc_float: true
    number_sections: false
    code_folding: hide
---  

<style type="text/css">  
/* Note: CSS uses C-style commenting. */
h1.title{font-size:22px; text-align:center;}
h4.author{font-size:16px; text-align:center;}
h4.date{font-size:16px; text-align:center;}
body{ /* Normal  */ font-size: 13px}
td {  /* Table   */ font-size: 12px}
h1 { /* Header 1 */ font-size: 16px}
h2 { /* Header 2 */ font-size: 14px}
h3 { /* Header 3 */ font-size: 12px}
.math{ font-size: 10pt;}
.hi{ /* hanging indents */ 
    padding-left:22px; 
    text-indent:-22px;
}
blockquote {  
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 12px;
    border-left: 5px solid #eee;
}
code.r{ /* code */ 
       font-size: 12px;
}
pre{/*preformatted text*/ 
    font-size: 12px;
}
p.caption {/* figure captions */ 
    font-size: 1.0em;
    font-style: italic; 
} 
</style>

```{r setup, echo=FALSE}
rm(list=ls()) # clean up
library(knitr)
library(coda)
gr <- (1+sqrt(5))/2 # golden ratio, for figures
opts_chunk$set(comment="  ",
               eval.after="fig.cap",
               collapse=TRUE, 
               dev="png",
               fig.width=7,
               out.width="95%",
               fig.asp=1/gr,
               fig.align="center",
               cache=TRUE,
               autodep=TRUE
               )
source("sourceMe_A7.R")
```  

# Introduction  

In Chapter 7 we move from basic Bayesian concepts to the foundations of Markov Chain Monte Carlo (MCMC). Pages 154—156 are optional but I hope you will study them anyway, as they are one of the most accessible treatments I have seen. In them Kruschke shows that the target PDF is an equilibrium point of the Metropolis procedure. His demonstration does _not_ show that it is a _stable_ equilibrium; in theory it could be an _unstable_ equilibrium point, like a ball perched on a mountain top, but in practice such problems are vanishingly rare.  

**Note** that this assignment comes with a file `sourceMe_A7.R` containing Kruschke's `HDIofMCMC()`. The file is sourced in the setup chunk; just put it in the same folder as your A7. I am not asking you to understand how `HDIofMCMC()` works, but try to do so anyway; it's a useful function that you can apply to any set of samples.  

# Tips for figures    
Here are a couple of tricks I sometimes use when making plots. Neither is essential.   

## Captions  
To make a caption with inserted values, put `eval.after="fig.cap"` (notice the quotes) and `fig.cap=fig.cap` (notice the lack of quotes) in the chunk header, and then create `fig.cap` as a string at the end of the chunk using `paste()`. The alternative is to use the narrative to make your caption with inline code for the variable values. I often make `eval.after="fig.cap"` a default chunk option in my setup chunk.    

## legend( )  
To pack extra information into a legend, use its `title` argument. Create it using `paste()` or `paste0()`, and embed `\n` to make it multiline. When you play these kind of games, the box around the legend can cause trouble; to get rid of the box put `bty="n"` in your `legend()` call. Using `legend()` like this isn't especially pretty, but it is quicker to code than annotating with `text()`.      

```{r experiments, eval=FALSE, include=FALSE}
## Find a bimodal likelihood for this exercise
f1 <- function(x) {
  c1 <- 0.7
  c2 <- 1 - c1
  a1 <- 10
  b1 <- 20
  a2 <- 20
  b2 <- 10
  print(c(c1/beta(a1,b1), c2/beta(a2,b2)))
  7*beta(a1,b1)*dbeta(x, a1, b1) + 3*beta(a2,b2)*dbeta(x, a2, b2)
}
plot(f1, xlim=c(0,1), ylab="f(x)")
```   

# Exercise 7.4  
(Not in text) (25 points)  
In this exercise you write a Metropolis algorithm to solve a 1-D problem that is slightly more realistic than the island-hopping politician considered in Section 7.2 of our text. (Using MCMC on any 1-D problem is like using a twelve pound hammer to kill a mosquito, but we overlook that here.) Suppose we have some [compositional data](https://en.wikipedia.org/wiki/Compositional_data) with just two possible components, A and B say, indexed by $\theta$. For example, if $\theta=0.9$ the mixture is 90% A and 10% B. Suppose our likelihood function is  

$$
L(\theta\,\vert D) \propto \tag{1}
7\,\theta^9(1-\theta)^{19} + \ 3\,\theta^{19}(1-\theta)^9
$$
and our prior is the Haldane prior $\theta^{-1}(1-\theta)^{-1}$. (a) Use the Metropolis algorithm to generate 2,000 samples of $\theta$ from the posterior and (b) make a histogram. (c) Use `coda::effectiveSize()` to calculate the effective size of your 2,000 samples. For the proposal distribution use the normal with SD = 0.2. (d) Add the posterior PDF to your histogram as a red line. (f) Use your samples to estimate $\Pr(0.2 \le \theta \le 0.5)$. (g) Use your samples to estimate $\Pr(0.3 \le \theta \le 0.4 \, \big\vert\, 0.1 \le \theta \le 0.8)$.

Don't forget to give your plot a legend and a caption. In lieu of a legend, you may annotate your plot roughly like Kruschke does in the panels of Figure 7.1, page 146.  

# Solution 7.4  
(Put your plot and narrative here.)  

# Exercise 7.5  
(Not in text) (25 points)  
Do the same problem you did in Ex 7.4, but use the Gibbs sampler instead of the Metropolis algorithm. 

# Solution 7.5  
(Put your plot and narrative here.)  

# Exercise 7.6  
(Not in text) (25 points)

## Preamble  
The two-coin example of this chapter is somewhat artificial, because the two coins are _independent_. In this exercise we estimate the difference in bias between two coins using MCMC, but without resort to either Metropolis or the Gibbs sampler. As you saw in class, deriving a useful mathematical formula for the PDF of a difference $\theta_1-\theta_2$ is not an easy task. However, it is easy to _sample_ the difference, and then use those samples as a proxy for the PDF itself. 

## The task  
You have two coins that cannot communicate with each other. You flip the first one $N_1=50$ times obtaining $z_1=35$ heads. You flip the second coin $N_2=40$ times, obtaining $z_1=18$ heads. Generate 2,000 samples of the difference $\phi=\theta_1-\theta_2$ and use them to make a histogram. Estimate (a) the expected value of $\phi$, (b) the SD of $\phi$ and (c) a 95% HDI for $\phi$. (d) Use your samples to estimate $\Pr(-0.1 \lt \phi \lt 0.1)$. Annotate your plot with the sample mean and 95% HDI the way Kruschke does, or use `legend()`.    

# Solution 7.6   
(Put your plot and narrative here.)  